Recognizing Strong Gravitational Lenses

### 2015-03-15 doc/Course_Paper/paper.tex
updating paper

%%%%%%%%% ABSTRACT
\begin{abstract}
    The detection of a large, representative set of strong gravitational lenses
    could greatly aid in our understanding of cosmology. Unfortunately they are
    quite rare, and the best techniques now revolve around squads of scientists
    manually scanning through images. This is presently borderline
    unsustainable and will be laughably inefficient with the advent of the
    Large Synoptic Survey Telescope. Here we examine the effectiveness of
    convolution neural networks and transfer learning for automated detection
    algorithms of strong gravitational lenses. We use images from the
    \textsc{Space Warps} project, a citizen science initiative to examine tens
    of thousands of fields of galaxies for the presence of strong gravitational
    lenses. 
    % TODO: Insert results!
    Scripts used in the analysis of this paper are freely available at
    \url{https://github.com/cpadavis/strongcnn}. The images are currently only
    available to those who contact the author, but will be available to the
    public in the near future.
\end{abstract}

@@

A consequence of Einstein's Theory of General Relativity is that mass bends the
path of light. Most of the time the deflections are very small; the original
'gravitational lens' that tested the veracity of Einstein's theory in the years
after the first world war was the sun, which deflected light from stars behind
it only a few seconds of arc. However, when light passes through a particularly
deep gravitational potential (say, near the center of the dark matter halo of a
galaxy cluster), the deflections can be particularly large, resulting in
brilliant arcs and multiple images. These strong deflections due to light
passing through a deep gravitational potential are termed strong gravitational
lenses. The gravitational lensing signal can heuristically be thought of as a
trade off between a couple factors: larger gradients in the gravitational
potential create larger distortions (and larger gradients in the gravitational
potential tend to reside near to the center of the foreground galaxy or galaxy
cluster); the more separated the foreground and background objects are, the
bigger the proportional distortion on the background object by the foreground
(but the farther away the background object is, the smaller it
appears\footnote{Note that this is only true in the ``low redshift'' universe:
  when objects are farther away than a cosmological redshift of $z \approx 2.5$
or about 2.5 Gyr after the birth of the Universe, they will actually
\textit{grow} in angular extent.}).

The very existence of these potentials acts as a verification of the Theory of
General Relativity, but they can also be used for much more. Strong
gravitational lenses are one of the few ways to directly probe the distribution
of dark matter, a particle (or possibly family of particles) that does not emit
electromagnetic radiation but does have mass and hence interacts
gravitationally with normal baryonic matter. This allows us to tally the mass
of the largest gravitationally-bound structures in the universe, galaxy
clusters, which can give us insight into the formation history of these massive
objects. In this way, strong gravitational lenses can then tell us something
about the expansion history of the universe, by setting limits on how massive
the most massive objects in the universe can be. The properties of the bent
light itself can also say much about that expansion history. When an object is
strongly-lensed into multiple images, each image travels a different span of
space and time. When an object does not vary much with time, these different
path lengths have no practical import. However, if the object varies
appreciably quickly (say it is a distant supermassive black hole at the center
of a galaxy whose accretion disk emits high-energy radiation at varying rates)
then these different path lengths can be used to pin down the rate of expansion
of the universe.

Unfortunately, for how useful strong gravitational lenses are, they are also
extremely rare. A next generation optical survey like the Large Synoptic Survey
Telescope can expect to find only ten thousand lenses in the whole sky, while
it will find ten billion galaxies. Currently in astronomy there are only order
hundreds of strong gravitational lenses known, mostly discovered by 'eyeball
squads' of graduate students. The small number means that target criteria must
be somewhat broad in order to maintain a relatively high completeness. Using
reasonable target criteria to find strong lenses such as looking only at
massive galaxies still means that nearly ten million objects will need to be
inspected in the next generation in order to find those ten thousand lenses. A
team of ten graduate students could expect to spend about 14 years looking at
these objects. Computer algorithms are not much better: most current machine
learning algorithms are woefully-underpowered for this task, and generally have
poor completeness or poor purity -- and often poor both. Additionally, some
algorithms are better at finding some types of lenses than others; some
perform well on the brilliant arcs, but poorly on the multiply-imaged objects,
or vice versa. New algorithms need to be developed to find more strong
gravitational lenses, and more strong lenses need to be found to power these
algorithms.

@@

\begin{figure*}%[!ht] 
\begin{minipage}[b]{0.48\linewidth}
\centering\epsfig{file=Figures/roc_curve_stage_1.pdf,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.48\linewidth}
\centering\epsfig{file=Figures/roc_curve_stage_2.pdf,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
  \caption{Receiver Operating Curves the \textsc{Space Warps} system and
    different linear classifiers trained on feature vectors extracted from a
    convolution neural network originally used to determine galaxy
  morphologies. We find that of the linear classifiers on the feature vectors,
support vector machine classifications perform best on the test dataset,
however all the feature vectors perform worse than the users themselves. Note
that the $x$-axis stops at a false positive rate of 0.5, and the $y$-axis
begins at a true positive rate of 0.5.}
\label{fig:roc_tl} 
\end{figure*}

% TODO: Redescribe the network, including data augmentation and such.
Transfer Learning provides a mechanism by which

@@

% TODO: Remake with linear classifiers.
% TODO: Include results from CNN as well.
% TODO: Touch up description
\begin{figure}%[!ht]
\centering
\epsfig{file=Figures/roc_curve.pdf,width=\linewidth,angle=0,clip=}
\caption{Receiver Operating Curves for the Random
  Forest Transfer Learning and the \textsc{Space Warps} systems, divided by
  stage.}
  \label{fig:roc}
\end{figure}

% TODO: make figure with probability distributions between classifiers and sw

@@

% Discuss how well things went. Or didn't go.

@@

%-------------------------------------------------------------------------
\section{Intermediate/Preliminary Results}

We have created the cutouts of our images from the catalogs.
We have begun constructing our own convolutional network with two
convolutional/max-pool layers and two fully-connected layers. Although we have
not yet trained on our images, we can load these images and run them
through our initialized classifier.

>>>>

%%%%%%%%% ABSTRACT
\begin{abstract}
    The detection of a large, representative set of strong gravitational lenses
    could greatly aid in our understanding of cosmology. Unfortunately they are
    quite rare, and the best techniques now revolve around squads of scientists
    manually scanning through images. This is presently borderline
    unsustainable and will be laughably inefficient with the advent of the
    Large Synoptic Survey Telescope. Here we examine the effectiveness of
    convolution neural networks and transfer learning for automated detection
    algorithms of strong gravitational lenses. We use images from the
    \textsc{Space Warps} project, a citizen science initiative to examine tens
    of thousands of fields of galaxies for the presence of strong gravitational
    lenses. 
    % TODO: Insert results!
    We find that using a convolution neural network trained on Galaxy
    morphologies as a feature extractor performs admirably but markedly worse
    than the citizen scientists.
    Scripts used in the analysis of this paper are freely available at
    \url{https://github.com/cpadavis/strongcnn}. The images are currently only
    available to those who contact the author, but will be available to the
    public in the near future.
\end{abstract}

@@

A consequence of Einstein's Theory of General Relativity
is that mass bends the path of light.\cite{Einstein:2007aa} Most of the time the deflections are very
small; the original 'gravitational lens' that tested the veracity of Einstein's
theory in the years after the first world war was the sun, which deflected
light from stars behind it only a few seconds of arc. However, when light
passes through a particularly deep gravitational potential (say, near the
center of the dark matter halo of a galaxy cluster), the deflections can be
particularly large, resulting in brilliant arcs and multiple images. These
strong deflections due to light passing through a deep gravitational potential
are termed strong gravitational lenses. The gravitational lensing signal can
heuristically be thought of as a trade off between a couple factors: larger
gradients in the gravitational potential create larger distortions (and larger
gradients in the gravitational potential tend to reside near to the center of
the foreground galaxy or galaxy cluster); the more separated the foreground and
background objects are, the bigger the proportional distortion on the
background object by the foreground (but the farther away the background object
is, the smaller it appears\footnote{Note that this is only true in the ``low
  redshift'' universe: when objects are farther away than a cosmological
  redshift of $z \approx 2.5$ or about 2.5 Gyr after the birth of the Universe,
they will actually \textit{grow} in angular extent.}).\cite{Kochanek:2006oz}

The very existence of these potentials acts as a verification of the Theory of
General Relativity, but they can also be used for much more. Strong
gravitational lenses are one of the few ways to directly probe the distribution
of dark matter, a particle (or possibly family of particles) that does not emit
electromagnetic radiation but does have mass and hence interacts
gravitationally with normal baryonic matter.\cite{Refsdal:1964aa} This allows us to tally the mass
of the largest gravitationally-bound structures in the universe, galaxy
clusters, which can give us insight into the formation history of these massive
objects. In this way, strong gravitational lenses can then tell us something
about the expansion history of the universe, by setting limits on how massive
the most massive objects in the universe can be. The properties of the bent
light itself can also say much about that expansion
history.\cite{Refsdal:1964ab, Kundic:1997aa, Barnacka:2014aa} When an object is
strongly-lensed into multiple images, each image travels a different span of
space and time. When an object does not vary much with time, these different
path lengths have no practical import. However, if the object varies
appreciably quickly (say it is a distant supermassive black hole at the center
of a galaxy whose accretion disk emits high-energy radiation at varying rates)
then these different path lengths can be used to pin down the rate of expansion
of the universe.

Unfortunately, for how useful strong gravitational lenses are, they are also
extremely rare. A next generation optical survey like the Large Synoptic Survey
Telescope can expect to find only ten thousand lenses in the whole sky, while
it will find ten billion galaxies. Currently in astronomy there are only order
hundreds of strong gravitational lenses known, mostly discovered by 'eyeball
squads' of graduate students. The small number means that target criteria must
be somewhat broad in order to maintain a relatively high completeness. Using
reasonable target criteria to find strong lenses such as looking only at
massive galaxies still means that nearly ten million objects will need to be
inspected in the next generation in order to find those ten thousand lenses. A
team of ten graduate students could expect to spend about 14 years looking at
these objects. Computer algorithms are not much better: most current machine
learning algorithms are woefully-underpowered for this task, and generally have
poor completeness or poor purity -- and often poor both. Additionally, some
algorithms are better at finding some types of lenses than others; some perform
well on the brilliant arcs, but poorly on the multiply-imaged objects, or vice
versa. For example, \cite{Kubo:2008aa} attempt to fit arc-like features in
images in order to find strong gravitational lenses, but this means that
multiply-imaged quasars are completely ignored. \cite{Agnello:2015aa} and
\cite{Chan:2014aa} in contrast develop an algorithm for finding gravitationally
lensed quasars based on catalog-level colors and magnitudes, precluding their
algorithm finding strong gravitational lens arcs. New algorithms need to be
developed to find more strong gravitational lenses, and more strong lenses need
to be found to power these algorithms. These algorithms need to not only
identify potential lenses accurately, but be able to make strong statements
about their contamination rates, as spectroscopic follow-up can be an expensive
endeavor. \cite{Momcheva:2015aa} performed spectroscopic follow-up on 9768
galaxies, finding 28 new strong gravitational lens systems, but taking 40
nights of telescope time on expensive telescopes.

@@

\begin{figure*}%[!ht] 
\begin{minipage}[b]{0.48\linewidth}
\centering\epsfig{file=Figures/roc_curve_stage_1.pdf,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.48\linewidth}
\centering\epsfig{file=Figures/roc_curve_stage_2.pdf,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
  \caption{Receiver Operating Curves the \textsc{Space Warps} system and
    different linear classifiers trained on feature vectors extracted from a
    convolution neural network originally used to determine galaxy
  morphologies. We find that of the linear classifiers on the feature vectors,
support vector machine classifications perform best on the test dataset,
however all the feature vectors perform worse than the users themselves. Note
that the $x$-axis stops at a false positive rate of 0.5, and the $y$-axis
begins at a true positive rate of 0.5. Truly random guessing (which results in
a 1:1 relationship between the true positive rate and the false positive rate)
would not show up on this graph.}
\label{fig:roc_tl} 
\end{figure*}

Transfer Learning relies on the idea that convolution neural networks that
perform similar tasks pick up similar features, as well as the observation that
lower levels in neural networks tend to be quite generic in the features they
pick out. They provide an answer to the scenario when there are too few data to
effectively train a complex system like deep convolution neural networks: start
the training of your new system from the results of training a similar system.
For the scope of this project, we chose to examine the effects of transfer
learning from galaxy morphology to strong gravitational lens identification. In
both cases, an input image of a galaxy is fed into the network, and some
classification is read out. Additionally, both networks need to differentiate
shapes in the central regions of the galaxy image (for example to find bars in
spiral galaxies) from outer regions (spiral arms, arcs, multiply-lensed
systems). The Galaxy Zoo competition provides an ideal candidate for transfer
learning because of these facts, and also because the image quality is
comparable between the Sloan Digital Sky Survey (the telescope survey on which
the Galaxy Zoo images were based) and the CFHTLS survey. We have on hand a
convolution neural network trained to classify galaxy morphology provided by
Ryan Keisler \footnote{\url{rkeisler@stanford.edu}}. The architecture of that
network can be seen in Figure~\ref{fig:ryan_architecture}. We do the simplest
thing possible: we run the convolution neural network as a feature extractor,
and take images from the fully connected layer. Thus we transform a
$96\times96\times3$ image into a 500 feature vector. We then train these
feature vectors on various classifiers (Random Forest, Support Vector Machine,
Softmax) and evaluate results against a test set. We distinguish between stage
1 and stage 2 data because the simulated lenses changed between the two sets.
The code that produces the feature vectors also augments the data by
automatically producing feature vectors of flips and rotations of the input
images. This allows us to increase the size of our input dataset nearly
20-fold.

We train these datasets on three classifiers: Random Forests (which are an
ensemble of decision trees trained on the data), Softmax and linear Support
Vector Machines. We use stochastic gradient descent for the latter two
classifiers. We create a test dataset by randomly extracting 20 percent of the
dataset and setting it aside. We also ensure that any data augmentation stays
in the training or test sets. Our goal with all the above systems is not to
find the maximal accuracy, but to find some reasonable trade-off between the
true positive rate and the false positive rate: we want to find as many lenses
as we can, but we also know that confirmation of these lenses by spectroscopic
follow-up is an expensive endeavor such that we want to minimize the number of
non-lenses that make it into our candidate list. Because of this, any potential
candidate list we would make from any of our classifiers has a relatively hard
threshold at a false positive rate of 0.2.

The resultant Receiver Operating Curves can be observed in
Figure~\ref{fig:roc_tl}. In general we find that support vector machines
perform the best as a classifier on the feature vectors, but that the feature
vectors perform more poorly than the \textsc{Space Warps} users. We must caveat
though that \textsc{Space Warps} does not create a validation dataset against
which to test the performance of the system. Even if we don't perform quite as
well as the citizen scientists, we consider this a
promising baseline for future performance by transfer learning: we have not
even begun to consider potential performance gains by retraining the
convolution neural network on the \textsc{Space Warps} data.

@@

% Do we need this section?

@@

% Discuss how well things went. Or didn't go.
The need for new automated detection algorithms for finding strong
gravitational lenses will only become more pressing in the next decade, as it
becomes infeasible for scientists to scan images by eye for such systems. Using
the \textsc{Space Warps} dataset, we have examined how convolution neural
networks trained both on this particular dataset and on other datasets can
perform at the detection task.
% TODO: brief blurb about our CNN
We find that features extracted from a convolution neural network trained on
the classification of galaxy morphology (with a linear support vector
classifier on top for converting the feature vector to a binary ``lens'' and
``not-lens'') performs admirably but markedly worse than the citizen scientists
trained on the dataset. Further work examining improvements by retraining the
whole neural network could lead to a generic classification machine that takes
images from \textit{any} galaxy survey and states whether the image contains a
strong lens or not. Additionally there is much potential in both direct
convolution neural networks and transfer learning from other networks in
linking the classification outputs of the networks with the probability
estimates of the \textsc{Space Warps} system, which also examined several
thousands more ``unknown'' systems and could lead to more gravitational lenses
being identified.

@@

%-------------------------------------------------------------------------
% References
